---
title: "codyn: An R package for analysis of long-term ecological community data"
author: "Lauren M. Hallett"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{codyn: An R package for analysis of long-term ecological community data}
  %\VignetteEngine{knitr::rmarkdown}
  %\SweaveUTF8
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---
From Collins08:
Most well understood measures of community structure
(e.g., diversity indices, rank-abundance curves,
species–area curves) have advantages in some contexts
and limitations in others. Many such measures represent
a ‘‘snapshot in time’’ yet many ecological models are
dynamic and require insight into patterns and mechanisms
that change over time. As long-term data sets
increase in breadth, length, and scope, better visualization
tools and analytical methods will be needed to
quantify temporal dynamics in ecological communities
(MacNally 2007, Alonso et al. 2008).


Many measures of community structure (eg, diversity indices and rank-abundance curve)  represent “snapshots in time” – which poorly capture the temporal dynamics of ecological systems. As long-term datasets increase in scope and length, new analytical tools are developing to capture patterns of species interactions over time. We present a new R package, codyn, to help ecologists easily implement these metrics and gain broader insight into community dynamics. codyn functions are explicitly temporal and include metrics of community stability and species turnover; they can also be implemented over multiple spatial replicates. In addition, codyn contains vignettes which reproduce figures and methods from published papers. 


A growing number of syntheses studies have documented high turnover in communities with static species richness, highlighting the need to utilize turnover metrics in tandem with diversity metrics to understand ecological community dynamics. 


## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
