---
title: "Temporal diversity indices"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  
  
---

#Overview
Many measures of community structure, such as diversity indices and rank-abundance curves, represent “snapshots in time” – which poorly capture the temporal dynamics of ecological systems. For example, species richness (i.e., the number of species present) is a common metric to characterize ecological communities. However, species richness may be a poor indicator of community change if there is turnover in the identity but not the number of species present over time [@Collins2008; @cleland2013]. Similarly, species rank abundances can reorder through time, even when the pool of species present remains relatively stable. Within-year rank abundance curves fail to capture this dynamic, and other tools are needed to quantify within-community reordering without masking informative temporal shifts in internal dynamics [@Collins2008]. `codyn` includes three functions to characterize temporal shifts in species identity and rank abundances over time:

- **`turnover`** calculates total turnover as well as the proportion of species that either appear or disappear between timepoints.

- **`rankshift`** quantifies relative changes in species rank abundances by taking the sum difference of species ranks in consecutive timepoints. 
This metric goes hand-in-hand "rank clocks," a useful visualization tool of shifts in species ranks. 

- **`rate_change`** analyzes differences in species composition between samples at increasing time lags. It reflects the rate of directional change in community composition. 


```{r echo=FALSE}
library(knitr)
```

# Example dataset
`codyn` utilizes a subset of the data presented by Collins et al. [-@Collins2008]. The `KNZ_Collins` dataset includes plant composition from the Konza Prairie Long-term Ecological Research site in Manhattan, KS. It spans 18 years and includes data collected from two locations, one that is annually burned and one that is unburned. 

```{r echo=FALSE}
KNZ_Collins<-read.csv(system.file("extdata", "collins08.csv", package="codyn"), sep=",", header=TRUE)
names(KNZ_Collins)[1]="location"
kable(head(KNZ_Collins))
```

#Species turnover
##Total turnover
The function `turnover` calculates three metrics of species turnover: total turnover, appearances, and disappearances.

The default metric `total` refers to total turnover, which calculates the proportion of species that differ between timepoints as:

$$ total turnover = \frac{Species gained + Species lost}{Total species observed in both timepoints} $$

The metric was introduced by MacArthur and Wilson (1963) and modified by Diamond (1969) to be expressed as a proportion in order to compare turnover between sites that differ in species richness.

`turnover` requires a dataframe with columns for species, year and abundance, and includes an argument to specify a column for spatial replicates.
```{r results='asis'}
library(codyn)
KNZ_turnover<-turnover(KNZ_Collins, replicate="location", species="species", year="year", abundance="abundance")
```
```{r echo=FALSE}
    kable(head(KNZ_turnover))
```

If there are no replicates in the dataset, specify `replicate=NA`. 
```{r results='asis'}
KNZ_burned<-subset(KNZ_Collins, location=="annually burned")
KNZ_burned_turnover<-turnover(KNZ_burned, replicate=NA, species="species", year="year", abundance="abundance")
```
```{r echo=FALSE}
kable(head(KNZ_burned_turnover))
```


Note that if `replicate=NA` but the dataset in fact includes replicates, `turnover` will calculate across all species present in a given year (i.e., it will return a site-level measure of turnover). For example, specifying `replicate=NA` while using the entire `KNZ_Collins` dataset yields total turnover across the unburned and burned locations combined.

```{r results='asis'}
KNZ_turnover_agg<-turnover(KNZ_Collins, replicate=NA, species="species", year="year", abundance="abundance")
```
```{r echo=FALSE}
    kable(head(KNZ_turnover_agg))
```

##Appearances and disappearances

Total turnover incorporates both species appearances and disappearances, but sometimes it is useful to parse their relative contribution. For example, a timepoint in which many species appear may reflect a different ecological story than a timepoint in which many species drop from the system, but the total turnover in both scenarios may be similar. 

Specifying `metric="appearance"` will return the proportion of species that appeared relative to the total number of species observed in both timepoints. As before, spatial replicates can be specified with the `replicate` argument; setting `replicate=NA` will calculate across the full dataset.
```{r results='asis'}
KNZ_appearance<-turnover(KNZ_Collins, replicate="location", species="species", year="year", abundance="abundance", metric="appearance")
```


Similarly, specifying `metric="disappearance"` will return the proportion of species that disappeared relative to the total number of species observed in both timepoints. 
```{r results='asis'}
KNZ_disappearance<-turnover(KNZ_Collins, replicate="location", species="species", year="year", abundance="abundance", metric="disappearance")
```

##Turnover at Konza
Total turnover indicates that in general there was greater fluctuations in the species present in the unburned than burned location at Konza.

```{r echo=FALSE}
library(ggplot2)
ggplot(KNZ_turnover, aes(x=year, y=total, color=location))+geom_line()+ theme_bw() #+ theme(legend.position="none")
```

Parsing total turnover by appearances and disappearances indicates that in there were generally more appearances in the unburned than burned location. In addition, the strong divergence in total turnover between locations in 1995 was because the unburned location lost many species that year while the burned locations lost none.


```{r echo=FALSE}
ggplot(KNZ_appearance, aes(x=year, y=appearance, color=location))+geom_line()+ theme_bw() + theme(legend.position="none")
ggplot(KNZ_disappearance, aes(x=year, y=disappearance, color=location))+geom_line()+ theme_bw() + theme(legend.position="none")
```


#Species rankshifts
##Rank clocks

##Rank shifts
Mean rank shifts are the relative change in species rank abundances, which provides an indication of shifts in relative abundance over time. It is calculated as:

$$ MRS = {\sum_{i=1}^{N} (|R_i,t+1 - R_i,t|})/N $$

where $N$ is the number of species in common in both years, $t$ is the year, $R_i,t$ is the relative rank of species $i$ in year $t$.

In `codyn`, this metric is calculated with `meanrankshift` and requires a dataset with a year, species and abundance column; it also include an option replicate column to calculate the metric within multiple spatial replicates. 

```{r}
KNZ_rankshift <- meanrankshift(KNZ_Collins, replicate="location", year="year", species="species", abundance="abundance")
#Select the final year from the returned year_pair
KNZ_rankshift$year<-as.numeric(substr(KNZ_rankshift$year_pair, 6,9))
```
```{r echo=FALSE}
kable(head(KNZ_rankshift))
```



If`replicate=NA` but the dataset in fact includes replicates, `meanrankshifts` will be calculated using all species present in the datset. For KNZ_Collins, this would return meanrankshifts for the burned and unburned areas combined.
```{r}
KNZ_rankshift_agg <- meanrankshift(KNZ_Collins, replicate=NA, year="year", species="species", abundance="abundance")
#Select the final year from the returned year_pair
KNZ_rankshift$year<-as.numeric(substr(KNZ_rankshift$year_pair, 6,9))
```

##Rank shifts at Konza
Calculating mean rank shifts highlights the stability of communities diverged at Konza around 1992, with fewer rank shifts between species in the burned relative to the unburned area. 
```{r}
##plotting
ggplot(KNZ_rankshift, aes(year, MRS, color=location)) + geom_line() + theme_bw()+ theme(legend.position="none")
```


##Citations


