---
title: "Community stability metrics"
author: ""
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{Community stability metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\SweaveUTF8
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}

---

# Overview
Ecologists have long debated the relationship between species diversity and stability. One key question in this debate is how the individual components of a community (e.g., species in species-rich communities) affect the stability of aggregate properties of the whole community (e.g., biomass production). It is increasingly recognized that unstable species populations may still maintain stable community productivity if a decrease in one species is compensated by an increase in another. In time series analysis, this should be reflected by a pattern in which species negatively covarying or fluctuating asynchronously while total community stability remained relatively stable. `codyn` includes a function to characterize community stability, **`community_stability`**, and three metrics to characterize species covariance and asynchrony:

- **`variance_ratio`** characterizes species covariance [@schluter1984; @houlahan2007], and includes a null-modeling approach to test significance [@hallett2014].

- **`synchrony`** has two options. The first compares the variance of the aggregated community with the variance of individual components [@loreau2008]. The second compares the average correlation of each individual species with the rest of the aggregated community [@gross2014].


```{r message=FALSE}
library(codyn)
library(knitr)
library(dplyr)
```

# Example dataset
To illustrate each function, we use a dataset of plant composition from the Konza Prairie Long-term Ecological Research site in Manhattan, KS. The KNZ dataset spans 24 years and includes 20 replicate subplots.

```{r results='asis'}
KNZ <- read.csv(system.file("extdata", "knz_001d.csv", package="codyn"), sep=",", header=TRUE)
kable(head(KNZ))
```


# Community stability
The `community_stability` function aggregates species abundances within replicate and year, and used these values to calculate community stability as µ/σ (i.e., the inverse of the coefficient of variation) [@tilman1999]. It includes an argument to calculate community stabilty across multiple replicates, and returns a dataframe with the name of the replicate column and the stability value.
```{r results='asis'}
KNZ_stability<-community_stability(KNZ, replicate="subplot", year="year", abundance="abundance")
kable(head(KNZ_stability))
```

If `replicate=NA`, the function will aggregate all values within a year and return an integer value.
```{r results='asis'}
KNZ_A1<-subset(KNZ, subplot=="A_1")
KNZ_A1_stability<-community_stability(KNZ_A1, replicate=NA, year="year", abundance="abundance")
KNZ_A1_stability
```

# Species covariance
## Variance Ratio

###Calculating the variance ratio
The variance ratio was one of the first metrics to characterize patterns of species covariance [@schluter1984] and was used in an early synthesis paper of species covariance in long time series [@houlahan2007]. The metric compares the variance of the community ($C$) as a whole relative to the sum of the individual population ($x_i$) variances:


$$ VR = \frac{Var(C)}{\sum_{i}^{N} Var(x_i)} $$

where:

$$ Var(C)  = \sum_{i = 1}^{N} Var(x_i) + 2\left(\sum_{i = 1}^{N - 1} \sum_{j = i + 1}^{N} Covar(x_i, x_j)\right) $$ 


If species vary independently then the variance ratio will be close to 1. A variance ratio  < 1 indicates predominately negative species covariance, whereas a variance ratio > 1 indicates that species generally covary positivey. 

### Significance testing
The variance ratio remains widely used but has been subject to a number of criticisms. Importantly, early uses of the variance ratio either did not include significance tests, or tested significance by comparing observed values to those returned by scrambling each species' time series. Null models using fully-scrambled species time series can generate spurious null expectations of covariance because the process disrupts species autocorrelation. Phase-scrambling [@Grman2010] and a temporal modification of the torus-translation [@hallett2014; adapted from @harms2001] have been used to address this issue. 

`variance_ratio` uses the temporal torus-translation to conduct null modeling for significance tests. In this method a starting year is randomly selected for each species’ time series. This generates a null community matrix in which species abundances vary independently but within-species autocorrelation is maintained (for each species, the time series is disrupted only once). 

The default `variance_ratio` setting calculates a null variance ratio for each replicate in the dataset, averages these values, and repeats as many times as specified by `bootnumber`. This vector of averaged, null variance ratios is then sampled for lower and upper confidence intervals, which are returned along with the observed variance ratio. 
```{r results='asis'}
KNZ_varianceratio<-varianceratio(KNZ, replicate="subplot", species="species", year="year", abundance="abundance",  
                                 bootnumber=10)

kable(KNZ_varianceratio)
```


Alternatively, specifying `averagereps=FALSE` will create a vector of null variance ratios for each replicate in the dataset, and return the subsequent confidence intervals and observed variance ratios for each replicate.
```{r results='asis'}
KNZ_varianceratio_avgrep<-varianceratio(KNZ, replicate="subplot", species="species", year="year", abundance="abundance",  
                                 bootnumber=10, averagereps=FALSE)

kable(head(KNZ_varianceratio_avgrep))
```


To indicate that there are no replicates in the dataset specify `replicate=NA`. 
```{r results='asis'}
KNZ_A1<-subset(KNZ, subplot=="A_1")
KNZ_A1_varianceratio<-varianceratio(KNZ_A1, replicate=NA, species="species", year="year", abundance="abundance",  
                                 bootnumber=10)

kable(KNZ_A1_varianceratio)
```

#### General torus-translation function
`codyn` also includes the option to apply the temporal torus-translation for other test statistices. 
`temporal_torus_translation` - returns the full null set
`temporal_torus_translation_CI`

###Variance ratio and species richness
A second criticism of the variance ratio is that it is sensitive to species richness. This is a particular concern when the metric is used to compare communities that have different levels of species richness. The most conservative approach is to restrict use of the variance ratio to two-species communities [@hector2010]. Comparing the effect size of the observed versus null variance ratio can also account for richness differences between communities. Two alternative metrics that quantify species asynchrony have been developed in part to respond to this issue.

## Synchrony Option 1: "Loreau"
Loreau and de Mazancourt (2008) developed a metric of species synchrony that compares the variance of aggregated species abundances with the summed variance of individual species:

$$ Synchrony = \frac{{σ_(x_T)}^{2}}{({\sum_{i} σ_(x_i)})^{2}}$$

where:

$$ x_T(t) = {\sum_{i=1}^{N} x_i(t))} $$

This measure of synchrony is standardized between 0 (perfect asynchrony) and 1 (perfect synchrony). A virtue of this metric is that it can be applied across communities of variable species richness. It can also be applied not only to species abundance but also population size and per capita growth rate. However, unlike the variance ratio it does not lend itself to significance testing. In addition, it will return similar values for communities shaped by different processes -- for example, even if species vary independently, the synchrony metric will be affected by the number of species and individual species variances [@gross2014].

In `codyn`, this is the default metric for the `synchrony` function and can be easily calculated for each replicate in a dataset. 
```{r results='asis'}
KNZ_synchrony<-synchrony(KNZ, replicate="subplot", species="species", year="year", abundance="abundance", metric="Loreau")
kable(head(KNZ_synchrony))
```


If there are no replicates within the dataset, specify `replicate=NA`. 
```{r results='asis'}
KNZ_A1<-subset(KNZ, subplot=="A_1")
KNZ_A1_synchrony<-synchrony(KNZ_A1, replicate=NA, species="species", year="year", abundance="abundance", metric="Loreau")
KNZ_A1_synchrony
```


## Synchrony Option 2: "Gross"
Gross et al. 2014 developed a metric of synchrony that compares the average correlation of each individual species with the rest of the aggregated community:

$$ Synchrony = (1/N){{\sum_{i}Corr(x_i, \sum_{i≠j}{x_j})}}$$

This measure of synchrony is standardized from -1 (perfect asynchrony) to 1 (perfect synchrony) and is centered at 0 when species fluctuate independently. A virtue of this metric is it not sensitive to richness and has the potential for null-model significance testing. It may under-perform on short time series because it is based on correlation, and care should be taken when applying it to communities that contain very stable species (i.e., whose abundances do not change throughout the time series).


In `codyn`, this metric is calculated with the "Gross" option in the `synchrony` function and can be easily calculated for each replicate in a dataset.
```{r results='asis'}
KNZ_synchrony_2<-synchrony(KNZ, replicate="subplot", species="species", year="year", abundance="abundance", metric="Gross")
kable(head(KNZ_synchrony))
```


If there are no replicates within the dataset, specify `replicate=NA`. 
```{r results='asis'}
KNZ_A1<-subset(KNZ, subplot=="A_1")
KNZ_A1_synchrony_2<-synchrony(KNZ_A1, replicate=NA, species="species", year="year", abundance="abundance", metric="Gross")
KNZ_A1_synchrony
```



## Citations


