---
title: "The temporal variance ratio for time series data"
author: "Andrew MacDonald"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: biblio.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r echo=FALSE, message=FALSE}
library(knitr)
```
  

There are several ways to look at community data over time.

```{r results='asis'}
knz_001d <- read.csv(system.file("extdata", "knz_001d.csv", package="codyn"), sep=",", header=TRUE)
kable(head(knz_001d))
```

## The Variance Ratio

Community variance of total biomass will be

$$ Var(x_1 + x_2 + \dotsb + x_N)  = \sum_{i = 1}^{N} Var(x_i) + 2\left(\sum_{i = 1}^{N - 1} \sum_{j = i + 1}^{N} Covar(x_i, x_j)\right)$$ 

Which gives us the covariance ratio [@Grman2010]:

$$ vr = \frac{var\left(\sum_{i}^{N} x_i \right)}{\sum_{i}^{N} var(x_i)} $$

The variance ratio measures the amount of covariance between species.  If species are overall uncorrelated (i.e. $Covar(x_i, x_j) = 0$) then the ratio's value will be close to 1. If species have a negative covariance then the ratio will be < 1, and if species have a positive covariance then it's value is > 1.

## Caluculating the variance ratio
We have a function for calculating the variance ratio:



## citations


