---
title: "Publications using codyn"
bibliography: citing-pubs.bib
nocite: '@*'
---

```{r setup, include = FALSE}
bib <- readLines("citing-pubs.bib", encoding = "UTF-8")
header_value <- function(name) {
  sub(paste0("^% ", name, ": "), "", grep(paste0("^% ", name, ": "), bib, value = TRUE))
}
n_pubs <- sum(grepl("^@", bib))
retrieved <- format(as.Date(header_value("Retrieved")), "%B %d, %Y")
dois <- strsplit(header_value("Sources"), ", ")[[1]]
years <- as.integer(sub("^\\s*year = \\{(\\d{4})\\},?$", "\\1",
                        grep("^\\s*year = ", bib, value = TRUE)))
```

codyn has been cited by **`r n_pubs` publications**. The list is compiled
automatically from [OpenAlex](https://openalex.org) and includes every work that
cites one of the papers or software releases describing codyn:
`r paste0("[", dois, "](https://doi.org/", dois, ")", collapse = ", ")`.
It was last updated on `r retrieved` and is refreshed monthly.

OpenAlex does not index everything, so theses, reports and some journals may be
missing. If your work uses codyn and isn't listed, please
[open an issue](https://github.com/NCEAS/codyn/issues). The list is also
available as a
[BibTeX file](https://github.com/NCEAS/codyn/blob/main/vignettes/articles/citing-pubs.bib).

```{r by-year, echo = FALSE, fig.width = 7, fig.height = 3.5, fig.alt = "Bar chart of the number of publications citing codyn in each year."}
counts <- table(factor(years, levels = seq(min(years), max(years))))
par(mar = c(3, 4, 1, 1), las = 1)
barplot(counts, col = "#4a7ab5", border = NA, ylab = "Publications")
```

Publications citing codyn each year (`r format(Sys.Date(), "%Y")` is a partial year).

## Publications

::: {#refs}
:::
