---
title: "Publications using codyn"
---

```{r setup, include = FALSE}
bib <- readLines("citing-pubs.bib", encoding = "UTF-8")
header_value <- function(name) {
  sub(paste0("^% ", name, ": "), "", grep(paste0("^% ", name, ": "), bib, value = TRUE))
}
n_pubs <- sum(grepl("^@", bib))
retrieved <- format(as.Date(header_value("Retrieved")), "%B %d, %Y")
dois <- strsplit(header_value("Sources"), ", ")[[1]]

# Split the BibTeX file into one string per entry, with its publication year.
entry_start <- cumsum(grepl("^@", bib))
entries <- vapply(split(bib, entry_start)[-1], paste, "", collapse = "\n")
years <- as.integer(sub("(?s).*\\n\\s*year = \\{(\\d{4})\\}.*", "\\1", entries, perl = TRUE))

# Format a set of BibTeX entries as an HTML reference list using pandoc's
# citeproc, the same way a bibliography in an R Markdown document is formatted.
format_references <- function(entries) {
  bib_file <- tempfile(fileext = ".bib")
  md_file <- tempfile(fileext = ".md")
  html_file <- tempfile(fileext = ".html")
  writeLines(entries, bib_file, useBytes = TRUE)
  writeLines(c("---", "nocite: '@*'", "---"), md_file)
  rmarkdown::pandoc_convert(md_file, to = "html", from = "markdown",
                            output = html_file,
                            options = c("--citeproc", paste0("--bibliography=", bib_file)))
  html <- readLines(html_file, encoding = "UTF-8")
  sub('id="refs" ', "", html, fixed = TRUE)
}
```

codyn has been cited by **`r n_pubs` publications**. The list is compiled
automatically from [OpenAlex](https://openalex.org) and includes every work that
cites one of the papers or software releases describing codyn:
`r paste0("[", dois, "](https://doi.org/", dois, ")", collapse = ", ")`.
It was last updated on `r retrieved` and is refreshed monthly.

OpenAlex does not index everything, so theses, reports and some journals may be
missing. If your work uses codyn and isn't listed, please
[open an issue](https://github.com/NCEAS/codyn/issues). The list is also
available as a
[BibTeX file](https://github.com/NCEAS/codyn/blob/main/vignettes/articles/citing-pubs.bib).

```{r by-year, echo = FALSE, fig.width = 7, fig.height = 3.5, fig.alt = "Bar chart of the number of publications citing codyn in each year."}
counts <- table(factor(years, levels = seq(min(years), max(years))))
par(mar = c(3, 4, 1, 1), las = 1)
barplot(counts, col = "#4a7ab5", border = NA, ylab = "Publications")
```

Publications citing codyn each year (`r format(Sys.Date(), "%Y")` is a partial year).

## Publications

Listed by year, newest first, and alphabetically by author within each year.

```{r references, echo = FALSE, results = "asis"}
for (year in sort(unique(years), decreasing = TRUE)) {
  cat("\n### ", year, " {#year-", year, "}\n\n```{=html}\n", sep = "")
  cat(format_references(entries[years == year]), sep = "\n")
  cat("\n```\n")
}
```
